---
title: "Geographic Disparities in Access to Telemedicine Services in the Post-COVID-19 Era"
author: "Shirisha Biyyala"
date: " `r Sys.Date()` "
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

## 1. Libraries

```{r load-libraries}
library(readr)
library(tidyverse)
library(knitr)
library(htmltools)
library(janitor)
library(dplyr)
library(stringr)
library(ggplot2)
library(treemapify) 
```

All the required libraries are loaded for subsequent operations.

## 2. Datasets

```{r load-datasets}
# Load datasets
enrollment <- read_csv("../data/PPEF_Enrollment_Extract_2024.10.07.csv", show_col_types = FALSE)
location <- read_csv("../data/PPEF_Practice_Location_Extract_2024.10.07.csv", show_col_types = FALSE)
reassignment <- read_csv("../data/PPEF_Reassignment_Extract_2024.10.07.csv", show_col_types = FALSE)
secondary <- read_csv("../data/PPEF_Secondary_Specialty_Extract_2024.10.07.csv", show_col_types = FALSE)

head(enrollment)
head(location)
head(reassignment)
head(secondary)
```
These datasets are now loaded into R from the `data` folder in the repository. We can use them for exploration and further analysis.

```{r glimpse-data}
glimpse(enrollment)
glimpse(location)
glimpse(reassignment)
glimpse(secondary)
```
## 3. Data Cleaning

```{r col-names-cleaning}
# Clean column names
enrollment <- enrollment %>% clean_names()
location <- location %>% clean_names()
reassignment <- reassignment %>% clean_names()
secondary <- secondary %>% clean_names()
```

The column names of the `enrollment`, `location`, `reassignment`, and `secondary` datasets were standardized using `clean_names()` to ensure consistent formatting.


```{r duplicate-missing-data}
# Remove duplicates and missing values
enrollment <- enrollment %>% distinct() %>%  select(-org_name) %>%  drop_na()
# org_name has NA values
location <- location %>% distinct() %>% drop_na()
reassignment <- reassignment %>% distinct() %>% drop_na()
secondary <- secondary %>% distinct() %>% drop_na()
```

#### Enrollment Dataset:
- Duplicate rows were removed.
- The `org_name` column was dropped as it contained missing values and was deemed unnecessary for the analysis.
- Remaining rows with missing values were removed using `drop_na()`.

#### Location, Reassignment, and Secondary Datasets:
- Duplicates were removed.
- Rows with missing values were dropped.

```{r colnames-after-cleaning}
colnames(enrollment)
colnames(location)
colnames(reassignment)
colnames(secondary)
```

```{r data-type-conversion}

# Enrollment dataset
enrollment <- enrollment %>%
  mutate(
    npi = as.character(npi),  # Convert NPI to character
    provider_type_cd = as.factor(provider_type_cd),  # Convert to factor
    provider_type_desc = as.factor(provider_type_desc),  # Convert to factor
    state_cd = as.factor(state_cd),  # Convert to factor
    gndr_sw = factor(gndr_sw, levels = c("F", "M", "9"), labels = c("Female", "Male", "Unknown"))  # Gender
  )

# Location dataset
location <- location %>%
  mutate(
    state_cd = as.factor(state_cd),  # Convert to factor
    zip_code = str_sub(zip_cd, 1, 5)  # Extract first 5 digits of ZIP
  )

# Reassignment dataset
reassignment <- reassignment %>%
  rename(enrlmt_id = reasgn_bnft_enrlmt_id)  # Rename column for consistency

# Secondary dataset
secondary <- secondary %>%
  mutate(
    provider_type_cd = as.factor(provider_type_cd),  # Convert to factor
    provider_type_desc = as.factor(provider_type_desc)  # Convert to factor
  )
```

#### Enrollment Dataset:
- `npi` was converted to a character type for consistency as it contains numerical IDs that should not be treated mathematically.
- `provider_type_cd` and `provider_type_desc` were converted to factors to represent categorical data.
- `state_cd` was converted to a factor for consistency in state representations.
- `gndr_sw` was recoded as a factor with labeled levels ("Female", "Male", "Unknown").

#### Location Dataset:
- `state_cd` was converted to a factor for consistency.
- `zip_code` was extracted as the first five digits of the `zip_cd` column to standardize the format.

#### Reassignment Dataset:
- The column `reasgn_bnft_enrlmt_id` was renamed to `enrlmt_id` for consistency across datasets.

#### Secondary Dataset:
- `provider_type_cd` and `provider_type_desc` were converted to factors.


## 4. Data Preprocessing

```{r summary}
summary(enrollment)
summary(location)
summary(reassignment)
summary(secondary)
```


```{r merge-data}
# Merge enrollment data with practice location data
merged_data <- merge(enrollment, location, by = "enrlmt_id")

# Merge with reassignment data
merged_data <- merge(merged_data, reassignment, by = "enrlmt_id")

# Merge with secondary specialty data
merged_data <- merge(merged_data, secondary, by = "enrlmt_id")

head(merged_data)
glimpse(merged_data)
summary(merged_data)
```

#### Enrollment and Location:
- Merged on the common key `enrlmt_id`.

#### Adding Reassignment Data:
- Merged the resulting dataset with the reassignment data using `enrlmt_id`.

#### Adding Secondary Specialty Data:
- Finally, merged the previous dataset with the secondary dataset using `enrlmt_id`.

```{r save-file}
# Save the final merged data as a CSV file
write.csv(merged_data, "../data/Medicare_Public_Provider_Enrollment_Data.csv", row.names = FALSE)
```

```{r data-to-consider}
# Select the relevant columns and rename them
df <- merged_data %>%
  select(
    enrlmt_id, npi, pecos_asct_cntl_id, provider_type_desc.x, state_cd.x, first_name, mdl_name, last_name, gndr_sw, city_name, zip_code, rcv_bnft_enrlmt_id, provider_type_desc.y, state_cd.y
  ) %>%
  rename(
    provider_type_desc.before = provider_type_desc.x,
    provider_type_desc.after = provider_type_desc.y,
    state_cd.before = state_cd.x,
    state_cd.after = state_cd.y
  )

# Check the first few rows of the new 'data'
head(df)
glimpse(df)

# Save the final  data as a CSV file
write.csv(df, "../data/Medicare_Provider_Data.csv", row.names = FALSE)
```

## Summary of Resulting Dataset
The merged dataset contains 14 columns with cleaned and transformed data. Key columns include:

- `provider_type_cd.before`, `provider_type_desc.before`, and `state_cd.before` (from the enrollment data).
- `city_name`, `zip_code`, and `state_cd.after` (from the location data).
- `rcv_bnft_enrlmt_id` (from the reassignment data).
- `provider_type_cd.after` and `provider_type_desc.after` (from the secondary data).

Rows with inconsistent or missing data were removed to ensure high-quality and complete information.
